Explaining thermal performance in fish: interplay of ecology and physiology drive growth and fitness response

### 2018-07-03 draft_model_only.Rnw
forward

Increasing temperatures under climate change are thought to affect individual physiology of ectotherms through increases in metabolic demands, leading to changes in species performance with concomitant effects on species ecology. Although intuitively appealing, the driving mechanisms behind thermal performance is contested: thermal performance (e.g., growth) appears correlated with metabolic scope (i.e., oxygen availability for activity) for a number of species, but a substantial number of datasets do not support oxygen limitation of long-term performance. Whether or not oxygen limitations via the metabolic scope, or a lack thereof, have major ecological consequences remains a highly contested question. Here, we propose a general size and trait-based model of energy and oxygen budgets to determine the relative influence of thermal sensitivity of metabolic rates, oxygen limitation, and environmental conditions on thermal performance. We show that oxygen limitation is not necessary to explain performance curves that are correlated with metabolic scope. Oxygen can drastically limit performance and fitness at temperature extremes, but changes in thermal performance are primarily driven by the interplay of changing metabolic rates and species ecology. Furthermore, our model reveals that fitness trends with temperature likely oppose trends in growth and efficiency, suggesting a potential explanation for the paradox that species often occur at lower temperatures than their growth-optimum. We suggest that this general model provides a mechanistic underpinning that can provide general and realistic predictions about temperature impacts on ectotherm performance and effects of climate change on species with different metabolic and ecological traits.

@@

\subsection{Model description}
Ectotherms adjust the relative amounts of time ($\tau$) spent on metabolically costly activity and resting/hiding to optimise the net energy gain relative to mortality. In the following, we refer to $\tau$ as the activity fraction for sake of generality.   Since both energy gain and metabolic losses are sensitive to temperature and oxygen limitations, both the activity level and the net energy gain will be subject to these environmental constraints. Their interplay thus determines available energy for growth and reproduction.

Net energy gain $P$ (mass per time) is the difference between supply $S$ and metabolic demands $D$, each being functions of body weight $w$ and temperature $T$:
\begin{align}
P(w,T) &= S(w,T) - D(w,T) \\
          &=(1-\beta-\phi)f(w,T) h c(T) w^q  - c(T) k w^n - \tau \delta c(T) k w  \\
  \label{eq:P}
\end{align}
where the feeding level ($[0:1]$)is given by a Holling type II function response:
\begin{align}
  f(w,T) &= \frac{\tau \gamma\Theta w^{p} }{\tau \gamma\Theta w^{p} + h c(T) w^q} }, \label{eq:f} \\
\end{align}
where $h c(T) w^q$ is the maximum consumption rate. The corresponding function for oxygen is given in Appendix A.  The feeding level ([0,1]) is determined by the fraction of time spent foraging $\tau$, foraging rate $\gamma w^p \Theta$ (search rate $\gamma w^p$ times prey availability $\Theta$) and maximum consumption $h c(T) w^q$.  The supply is discounted by the loss due to specific dynamic action  $\beta$ (SDA, or heat increment; the energy spent absorbing food), and $\phi$ is the fraction of food excreted and egested.

@@

Fish will adjust their activity fraction $\tau$ to maximise fitness. We use Gilliam's rule as a fitness proxy \cite{gilliam_habitat_1987}:
\begin{equation}
  \tau^* = \argmax_{\tau} \{ \frac{P(w,T)}{M(w)} \}.
\end{equation}
This optimisation represents a "short-sighted" fitness optimisation that does not account for future changes in conditions, and is appropriate for optimisation in stable environments (\cite{sainmont_effective_2015}).  Mortality scales with activity fraction and weight as $w^{q-1}$ (\cite{andersen_how_2009, hartvig_food_2011}):
\begin{equation}
  \label{eq:M}
  M(w) = (\rho+\eta\tau) w^{q-1},
\end{equation}
$\rho$ is mortality at mass $w=1$ and $\tau=0$, that is, with no activity beyond that covered by standard metabolism.  Activity is limited by available oxygen, such that $P_{O_2}$ is not allowed to be negative \emph{Is that a correct interpretation?}.

>>>>

Increasing temperatures under climate change are thought to affect individual physiology of ectotherms through increases in metabolic demands, leading to changes in species performance with concomitant effects on species ecology. Although intuitively appealing, the driving mechanism behind thermal performance is contested: thermal performance (e.g., growth) appears correlated with metabolic scope (i.e., oxygen availability for activity) for a number of species, but a substantial number of datasets do not support oxygen limitation of long-term performance. Whether or not oxygen limitations via the metabolic scope, or a lack thereof, have major ecological consequences remains a highly contested question. Here, we propose a general size and trait-based model of energy and oxygen budgets to determine the relative influence of thermal sensitivity of metabolic rates, oxygen limitation, and environmental conditions on fish thermal performance. We show that oxygen limitation is not necessary to explain performance curves that are correlated with metabolic scope. Oxygen can drastically limit performance and fitness, especially at temperature extremes, but changes in thermal performance are primarily driven by the interplay of changing metabolic rates and species ecology. Furthermore, our model reveals that fitness trends with temperature can oppose trends in growth, suggesting a potential explanation for the paradox that species often occur at lower temperatures than their growth-optimum. We suggest that this general model provides a mechanistic underpinning that can provide general and realistic predictions about temperature impacts on the performance of fish and other ectotherms, and act as a null model for temperature impacts on species with different metabolic and ecological traits.

@@

\subsection{Model description}
Ectotherms adjust the relative amounts of time ($\tau$) spent on metabolically costly activity and resting/hiding to optimise the net energy gain relative to mortality (\cite{gilliam_habitat_1987}). In the following, we refer to $\tau$ as the activity fraction for sake of generality.   Since both energy gain and metabolic losses are sensitive to temperature and oxygen limitations, both the activity level and the net energy gain will be subject to these environmental constraints. Their interplay thus determines available energy for growth and reproduction.

Net energy gain $P$ (mass per time) is the difference between supply $S$ and metabolic demands $D$, each being functions of body weight $w$ and temperature $T$:
\begin{align}
P(w,T) &= S(w,T) - D(w,T) \\
          &=(1-\beta-\phi)f(w,T) h c(T) w^q  - c(T) k w^n - \tau \delta c(T) k w  \\
  \label{eq:P}
\end{align}
where the feeding level ($[0:1]$)is given by a Holling type II function response:
\begin{align}
  f(w,T) &= \frac{\tau \gamma\Theta w^{p} }{\tau \gamma\Theta w^{p} + h c(T) w^q}, \label{eq:f} \\
\end{align}
where $h c(T) w^q$ is the maximum consumption rate. The corresponding function for oxygen is given in Appendix A.  The feeding level ([0,1]) is determined by the fraction of time spent foraging $\tau$, foraging rate $\gamma w^p \Theta$ (search rate $\gamma w^p$ times prey availability $\Theta$) and maximum consumption $h c(T) w^q$.  The supply is discounted by the loss due to specific dynamic action  $\beta$ (SDA, or heat increment; the energy spent absorbing food), and $\phi$ is the fraction of food excreted and egested.

@@

Fish will adjust their activity fraction $\tau$ to maximise fitness. We use Gilliam's rule as a fitness proxy \cite{gilliam_habitat_1987}:
\begin{equation}
  \tau^* = \text{argmax}_{\tau} \{ \frac{P(w,T)}{M(w)} \}.
\end{equation}
This optimisation represents a "short-sighted" fitness optimisation that does not account for future changes in conditions, and is appropriate for optimisation in stable environments (\cite{sainmont_effective_2015}).  Mortality scales with activity fraction and weight as $w^{q-1}$ (\cite{andersen_how_2009, hartvig_food_2011}):
\begin{equation}
  \label{eq:M}
  M(w) = (\rho+\eta\tau) w^{q-1},
\end{equation}
$\rho$ is mortality at mass $w=1$ and $\tau=0$, that is, with no activity beyond that covered by standard metabolism.  Activity is limited by available oxygen, such that $P_{O_2}$ is not allowed to be negative \emph{Is that a correct interpretation?}.
